Tree growth, mortality, and climatic water deficit in west-coast states, USA

### 2025-12-12 Manuscript.Rmd
Incorporated edits from B. Frank.  This added a figure, changed the supplimental and main RMarkdown files, and altered many of the scripts involved in generating plots and tables.

---
title: "Tree growth, mortality, and climatic stress in west-coast states, USA"
author:
  - name: "Jeremiah D. Groom"
    affiliations:
      - name: "Groom Analytics LLC"
        address: "1975 SE Crystal Lake Dr., Unit 173"
        city: "Corvallis"
        state: "OR"
        postal-code: "97333"
    orcid: "0000-0002-0110-1036"
    email: "jeremy@groomanalytics.com"
  - name: "Bryce Frank"
    affiliations:
      - name: "USDA Forest Service"
        department: "Pacific Northwest Field Station"
        address: "3625 93rd Ave SW"
        city: "Olympia"
        state: "WA"
        postal-code: "98512"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  officedown::rdocx_document:
    reference_docx: "template_use2.docx"

@@

```{r load-data}
# Some general analysis info + plots
analysis.stats <- read_rds(paste0(RESULTS.OTHER, "analysis.stats.RDS"))

@@

# Significant findings, Growth
grow.n.AvB <- nrow(grow.diff$AvB2)
grow.n.A <- nrow(grow.diff$A_LMH2)
grow.n.B <- nrow(grow.diff$B_LMH2)

grow.AvB.sig <- length(grow.diff$AvB2$significant[grow.diff$AvB2$significant == TRUE])
grow.AvB.signeg <- length(grow.diff$AvB2$significant[grow.diff$AvB2$significant == TRUE & grow.diff$AvB2$Means < 0])
grow.A.sig <- length(grow.diff$A_LMH2$significant[grow.diff$A_LMH2$significant == TRUE])
grow.A.signeg <- length(grow.diff$A_LMH2$significant[grow.diff$A_LMH2$significant == TRUE & grow.diff$A_LMH2$Means < 0])
grow.B.sig <- length(grow.diff$B_LMH2$significant[grow.diff$B_LMH2$significant == TRUE])
grow.B.signeg <- length(grow.diff$B_LMH2$significant[grow.diff$B_LMH2$significant == TRUE & grow.diff$B_LMH2$Means < 0])

grow.tot.sig <- grow.AvB.sig + grow.A.sig + grow.B.sig
grow.tot <- grow.n.AvB + grow.n.A + grow.n.B

# Significant findings, Mort
mort.n.AvB <- nrow(mort.diff$AvB2)
mort.n.A <- nrow(mort.diff$A_LMH2)
mort.n.B <- nrow(mort.diff$B_LMH2)

mort.AvB.sig <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE])
mort.AvB.signeg <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE & mort.diff$AvB2$Means < 0])
mort.A.sig <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE])
mort.A.signeg <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE & mort.diff$A_LMH2$Means < 0])
mort.B.sig <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE])
mort.B.signeg <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE & mort.diff$B_LMH2$Means < 0])

mort.tot.sig <- mort.AvB.sig + mort.A.sig + mort.B.sig
mort.tot <- mort.n.AvB + mort.n.A + mort.n.B

# Summary of significant Growth and Mortality findings for table creation.
datasets.grow <- list(grow.diff$AvB2, grow.diff$A_LMH2, grow.diff$B_LMH2)
datasets.mort <- list(mort.diff$AvB2, mort.diff$A_LMH2, mort.diff$B_LMH2)
datatypes <- list("Above vs. Below", "Above", "Below")
n.comp.grow <- list(grow.n.AvB, grow.n.A, grow.n.B)
n.comp.mort <- list(mort.n.AvB, mort.n.A, mort.n.B)

@@

```{r table-creation}
# State growth
grow.state.cm <- cm2.fcn(1, grow.state)
grow.all.cm <- cm2.fcn(1, grow.all)
s.g.dat <- state.table.fcn(grow.state.cm, grow.all.cm, 1)

@@

# State mortality
s.m.dat <- state.table.fcn(mort.state, mort.all, 2)

@@

# CWD amounts
min.mean.init <- analysis.stats$init.cwd.stats$Mean[analysis.stats$init.cwd.min.mean]
min.mean.init.spp <- analysis.stats$init.cwd.stats$Species[analysis.stats$init.cwd.min.mean]
max.mean.init <- analysis.stats$init.cwd.stats$Mean[analysis.stats$init.cwd.max.mean]
max.mean.init.spp <- analysis.stats$init.cwd.stats$Species[analysis.stats$init.cwd.max.mean]

@@

tab.tbl_growth <- tab_nums(name = "tbl_growth", 
                        caption = "Decadal growth estimates by state and overall for analysis species (cm\u00B2/decade). Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots.")

tab.tbl_mort <- tab_nums(name = "tbl_mort", 
                        caption = "Decadal mortality estimates by state and overall for analysis species. Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots.")

@@

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "This figure illustrates our bivariate domain approach using *Calocedrus decurrens* as an example. Here we depict decadal (A) growth and (B) mortality domain estimates.  The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'AL' is Above CWD change threshold, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line indicating the CWD change threshold and the orange vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

fig.diff_grow <- fig_nums(name = "fig_diffgrow", 
                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for Figure ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous. Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05, while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively."))

fig.diff_mort <- fig_nums(name = "fig_diffmort", 
                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for Figure ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are coniferous; those below are deciduous.  Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05 while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively. "))

@@

Western North American forests are experiencing accelerating tree mortality rates in response to intensifying drought and water stress driven by climate change.  Tree mortality and growth reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).  However, obtaining design-based estimates for these demographic rates requires addressing a methodological challenge: revisit data are necessary to measure tree fate and growth, but plots differ in their initial water stress conditions and the magnitude of change between visits, necessitating a bivariate rather than univariate estimation approach. We developed a bivariate design-based estimation framework that stratifies plots into domains defined by both baseline initial-visit climatic water deficit (CWD) and change in CWD between visits.  We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species over decade-long intervals.  We estimated decadal growth and mortality overall, by state, and across six CWD domains per species, using bootstrap confidence intervals and Monte Carlo permutation tests to assess statistical significance.  Species responses were highly variable, with no universal pattern of increased mortality or reduced growth in domains experiencing elevated water stress. Several species exhibited overall mortality rates exceeding 0.20 per decade (20% decadal mortality), with state estimates of 0.34 for *Picea engelmannii* in Washington, 0.27 for *Acer macrophyllum* in California, and 0.37 for *Pinus contorta* in Washington. Domain-specific patterns aligned with known regional disturbances (bark beetles, drought, fire) for multiple species. Our bivariate framework provides a template for monitoring forest vital rates as ecosystem-level indicators of climate change response, while our species-specific findings identify or validate populations warranting immediate conservation assessment and inform forest management strategies.

@@

Carbon starvation and hydraulic failure can drive drought-induced mortality (see reviews by Adams et al. 2010, Sala et al. 2010, 2012, and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., piñon pine, *Pinus edulus*; Breshears et al. 2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 26 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022).

Drought stress can increase tree susceptibility to wildfire, insects, and disease.  Wildfire area (Abatzoglou and Williams, 2016; Parks et al. 2025) and number (Westerling 2016) are positively associated with increasing aridity in the western US.   Bark beetle outbreaks in semi-arid areas of the western US may be positively affected by or associated with acute intense drought and water-stressed trees (Negrón 2009, Gaylord et al. 2015) but chronically stressed trees may not provide bark beetle populations with the resources sufficient to promote population expansion (Huberty and Denno, 2004).  Norway spruce (*Picea abies*) saplings inoculated with the necrotrophic pathogenic fungi *Heterobasidion annosum* sensu stricto and exposed to drought conditions were more likely to die than control saplings or saplings exposed to either drought conditions or inoculation, although Scots pine (*Pinus sylvestris*) seedlings did not show a similar relationship (Gomez-Gallego et al. 2022). Other insects and disease may be promoted or inhibited by drought, depending on the circumstances and due to conflicting factors (Kolb et al. 2016).

@@

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.   Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced probability sample (McRoberts 2005, Reams et al. 2005, Woudenberg et al. 2010).  The sample allows for relatively unbiased design-based inference of population characteristics across a large geographical region.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates. Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of FIA plot temperatures for seedlings from those of trees. FIA plot revisit data were mostly incomplete at the time of their analysis.  Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots that gained trees of a given species and plots that lost trees.

@@

Our study has two objectives.  The first is to demonstrate, across 23 species, a bivariate design-based estimation approach for tree growth and mortality that simultaneously accounts for initial climatic conditions and change between measurements.  This addresses a gap in design-based analyses of forest change.  Second, we provide current growth and mortality estimates for our analysis species across our study area and by state to establish species status and to inform our climate analysis findings.

@@

Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.  Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al. 2005).  We constrained our analysis to only include remeasurements for areas considered forestland during both visits.

FIA relies on three related sampling procedures.  Phase 1 involves the classification of distributed points of remote sensing data into (at least) the categories of forest and non-forest.  Phase 1 plots are useful for estimation stratification.  The phase 2 permanent ground plots are a subset of Phase 1 points.  They consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from microplots, where seedlings and saplings are measured, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees. Phase 3 plots are an intensification of Phase 2 plots; we do not rely on their data here. Reams et al. (2005) and Bechtold and Scott (2005) provide plot design details and measurement protocols.

The FIA program selected plots with a spatially-balanced probability sampling design (McRoberts 2005, Reams et al., 2005) that used a target plot density of one ground plot every 24 km$^2$. Some regions and ownerships intensify their plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine stratum weights.

Plots in the three states are sampled every 10 years with a target of annually sampling a panel of 10% of plots every year (Thompson 2015). Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure. Tree growth is the observed increase in marked tree bole diameter between visits (cm$^2$) while mortality is captured from the change in tree condition. We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

Individual trees within plots were retained in the analysis if they were alive at the initial visit and met the stem diameter criterion during that visit (i.e., plot ingrowth was not included).  The growth analysis additionally restricted trees considered to those that were alive at both visits.  The mortality analysis excluded trees killed through silvicultural practices.  Both analyses excluded trees that shrank in diameter, physically moved due to landslides or other event, were erroneously tallied at either visit, were missed due to data collection procedural changes, or existed at any time in a nonsampled area.  We used only FIA plots sampled specifically for tree mortality and growth.

@@

We obtained plot-level CWD data from the model TerraClimate (Abatzoglou et al. 2018) for June, July, and August (summer months) between 1980 and 2023. Latitude and longitude were based on FIA-fuzzed coordinates.  We defined the 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year (initial CWD = mean(summer CWD from visit year$_{-19}$ to year$_0$).  We then found the mean of summer CWD values between visits and including the second visit year (second visit CWD = mean(summer CWD from year$_1$ to year$_10$) .  We then defined delta CWD = second visit CWD - initial CWD.  Figure 1 presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.

## Analysis
We obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16 in Scott et al. 2005).  We expanded the number of sampled trees in each plot by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits. The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.  The denominator for mortality is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.

>>>>

---
title: "Tree growth, mortality, and climatic water deficit in west-coast states, USA"
author:
  - name: "Jeremiah D. Groom"
    affiliations:
      - name: "Groom Analytics LLC"
        address: "1975 SE Crystal Lake Dr., Unit 173"
        city: "Corvallis"
        state: "OR"
        postal-code: "97333"
    orcid: "0000-0002-0110-1036"
    email: "jeremy@groomanalytics.com"
  - name: "Bryce Frank"
    affiliations:
      - name: "USDA Forest Service"
        department: "Pacific Northwest Field Station"
        address: "3625 93rd Ave SW"
        city: "Olympia"
        state: "WA"
        postal-code: "98512"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  officedown::rdocx_document:
    reference_docx: "template_use2.docx"

@@

```{r load-data}
n.spp <- length(sel.spp)

# Some general analysis info + plots
analysis.stats <- read_rds(paste0(RESULTS.OTHER, "analysis.stats.RDS"))

@@

# Function to extract a value out of the contrast figures by species and contrast
fig.val.fcn <- function(dat, spp.x, cont, rnd){
dat.source <- if(dat == "g") grow.diff else mort.diff
k <- if(dat == "g") 1 else 2
IvS2 <- cm2.fcn(k, results.table = dat.source$IvS2)
I_LMH2 <- cm2.fcn(k, results.table = dat.source$I_LMH2)
S_LMH2 <- cm2.fcn(k, results.table = dat.source$S_LMH2)
tab.comb <- bind_rows(IvS2, I_LMH2, S_LMH2)

val.out <- round(tab.comb$Means[tab.comb$Species == spp.x & tab.comb$Domain == cont], rnd)
return(val.out)
}

# Significant findings, Growth
grow.n.IvS <- nrow(grow.diff$IvS2)
grow.n.I <- nrow(grow.diff$I_LMH2)
grow.n.S <- nrow(grow.diff$S_LMH2)

grow.IvS.sig <- length(grow.diff$IvS2$significant[grow.diff$IvS2$significant == TRUE])
grow.IvS.signeg <- length(grow.diff$IvS2$significant[grow.diff$IvS2$significant == TRUE & grow.diff$IvS2$Means < 0])
grow.I.sig <- length(grow.diff$I_LMH2$significant[grow.diff$I_LMH2$significant == TRUE])
grow.I.signeg <- length(grow.diff$I_LMH2$significant[grow.diff$I_LMH2$significant == TRUE & grow.diff$I_LMH2$Means < 0])
grow.S.sig <- length(grow.diff$S_LMH2$significant[grow.diff$S_LMH2$significant == TRUE])
grow.S.signeg <- length(grow.diff$S_LMH2$significant[grow.diff$S_LMH2$significant == TRUE & grow.diff$S_LMH2$Means < 0])

grow.tot.sig <- grow.IvS.sig + grow.I.sig + grow.S.sig
grow.tot <- grow.n.IvS + grow.n.I + grow.n.S

# Significant findings, Mort
mort.n.IvS <- nrow(mort.diff$IvS2)
mort.n.I <- nrow(mort.diff$I_LMH2)
mort.n.S <- nrow(mort.diff$S_LMH2)

mort.IvS.sig <- length(mort.diff$IvS2$significant[mort.diff$IvS2$significant == TRUE])
mort.IvS.signeg <- length(mort.diff$IvS2$significant[mort.diff$IvS2$significant == TRUE & mort.diff$IvS2$Means < 0])
mort.I.sig <- length(mort.diff$I_LMH2$significant[mort.diff$I_LMH2$significant == TRUE])
mort.I.signeg <- length(mort.diff$I_LMH2$significant[mort.diff$I_LMH2$significant == TRUE & mort.diff$I_LMH2$Means < 0])
mort.S.sig <- length(mort.diff$S_LMH2$significant[mort.diff$S_LMH2$significant == TRUE])
mort.S.signeg <- length(mort.diff$S_LMH2$significant[mort.diff$S_LMH2$significant == TRUE & mort.diff$S_LMH2$Means < 0])

mort.tot.sig <- mort.IvS.sig + mort.I.sig + mort.S.sig
mort.tot <- mort.n.IvS + mort.n.I + mort.n.S

# Summary of significant Growth and Mortality findings for table creation.
datasets.grow <- list(grow.diff$IvS2, grow.diff$I_LMH2, grow.diff$S_LMH2)
datasets.mort <- list(mort.diff$IvS2, mort.diff$I_LMH2, mort.diff$S_LMH2)
datatypes <- list("Increasing vs. Stable", "Increasing", "Stable")
n.comp.grow <- list(grow.n.IvS, grow.n.I, grow.n.S)
n.comp.mort <- list(mort.n.IvS, mort.n.I, mort.n.S)

@@

```{r table-creation}
# State growth
grow.state.cm <- cm2.fcn(1, grow.state)
grow.all.cm <- cm2.fcn(1, grow.all)
s.g.dat <- state.table.fcn(grow.state.cm, grow.all.cm, 1) %>%
  left_join(spp.cat %>% dplyr::select(-spp_cat), by = c(`Species Name` = "Species")) %>%
  rename("Category" = "cat_abbr") %>%
  relocate(Category, .after = `Species Name`)

@@

# State mortality
s.m.dat <- state.table.fcn(mort.state, mort.all, 2) %>%
  left_join(spp.cat %>% dplyr::select(-spp_cat), by = c(`Species Name` = "Species")) %>%
  rename("Category" = "cat_abbr") %>%
  relocate(Category, .after = `Species Name`)

@@

# CWD amounts, tapping values from the initial & delta CWD tables in the suppliment
init.cwd.stats <- analysis.stats$init.cwd.stats %>%
  mutate(range05.95 = q95 - q05) %>%
  arrange(Mean)

delta.cwd.stats <- analysis.stats$diff.cwd.stats %>%
  mutate(range05.95 = q95 - q05) %>%
  arrange(Mean)

cwd.summ.fcn <- function(data.2use, col.interest, place, stat.use) {
end.start <- switch(stat.use, "min" = 1, "max" = n.spp) # Which row end (top or bottom) to start on?
place2 <- if(stat.use == "min") place else n.spp - place + 1

x <- data.2use %>% arrange(get(col.interest))
x2 <- x[place2, ]
x2.range <- paste0("(5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
x2.range.full <- paste0("(5th-95th percentile range = ", x2$range05.95, "mm, 5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
x2.mean.full <- paste0("(mean = ", x2$Mean, "mm, 5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")

return(list(spp.out = x2$Species, range.alone = x2$range05.95, mean.alone = x2$Mean,
            spp.range = x2.range, 
            spp.range.full = x2.range.full, spp.mean.full = x2.mean.full
            ))
}

#Initial CWD: Tree species with the min/max means and ranges.
init.mean.sm <- cwd.summ.fcn(init.cwd.stats, col.interest = "Mean", place = 1, stat.use = "min") #smallest mean
init.mean.sm2 <-   cwd.summ.fcn(init.cwd.stats, col.interest = "Mean", place = 2, stat.use = "min") # 2nd smallest mean
init.range.sm <- cwd.summ.fcn(init.cwd.stats, col.interest = "range05.95", place = 1, stat.use = "min") # smallest range
init.range.sm2 <- cwd.summ.fcn(init.cwd.stats, col.interest = "range05.95", place = 2, stat.use = "min")
init.mean.lg <- cwd.summ.fcn(init.cwd.stats, col.interest = "Mean", place = 1, stat.use = "max") # largest mean
init.mean.lg2 <- cwd.summ.fcn(init.cwd.stats, col.interest = "Mean", place = 2, stat.use = "max")
init.range.lg <- cwd.summ.fcn(init.cwd.stats, col.interest = "range05.95", place = 1, stat.use = "max")
init.range.lg2 <- cwd.summ.fcn(init.cwd.stats, col.interest = "range05.95", place = 2, stat.use = "max")

#Initial CWD: Tree species with the min/max means and ranges.
delta.mean.sm <- cwd.summ.fcn(delta.cwd.stats, col.interest = "Mean", place = 1, stat.use = "min") #smallest mean
delta.mean.sm2 <-   cwd.summ.fcn(delta.cwd.stats, col.interest = "Mean", place = 2, stat.use = "min") # 2nd smallest mean
delta.range.sm <- cwd.summ.fcn(delta.cwd.stats, col.interest = "range05.95", place = 1, stat.use = "min") # smallest range
delta.range.sm2 <- cwd.summ.fcn(delta.cwd.stats, col.interest = "range05.95", place = 2, stat.use = "min")
delta.mean.lg <- cwd.summ.fcn(delta.cwd.stats, col.interest = "Mean", place = 1, stat.use = "max") # largest mean
delta.mean.lg2 <- cwd.summ.fcn(delta.cwd.stats, col.interest = "Mean", place = 2, stat.use = "max")
delta.range.lg <- cwd.summ.fcn(delta.cwd.stats, col.interest = "range05.95", place = 1, stat.use = "max")
delta.range.lg2 <- cwd.summ.fcn(delta.cwd.stats, col.interest = "range05.95", place = 2, stat.use = "max")

min.mean.init <- analysis.stats$init.cwd.stats$Mean[analysis.stats$init.cwd.min.mean]
min.mean.init.spp <- analysis.stats$init.cwd.stats$Species[analysis.stats$init.cwd.min.mean]
max.mean.init <- analysis.stats$init.cwd.stats$Mean[analysis.stats$init.cwd.max.mean]
max.mean.init.spp <- analysis.stats$init.cwd.stats$Species[analysis.stats$init.cwd.max.mean]

@@

tab.tbl_growth <- tab_nums(name = "tbl_growth", 
                        caption = "Decadal growth estimates by state and overall for analysis species (cm\u00B2/decade). Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots. Ecological category abbreviations (Category) are provided in the text.")

tab.tbl_mort <- tab_nums(name = "tbl_mort", 
                        caption = "Decadal mortality estimates by state and overall for analysis species. Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots. Ecological category abbreviations (Category) are provided in the text.")

@@

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "This figure illustrates our domain analysis approach using *Calocedrus decurrens* as an example. Here we depict decadal (A) growth and (B) mortality domain estimates.  The six domains are defined as FIA plots that fall into the [I]ncreasing or [S]table CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'IL' is in the Increasing CWD change category, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line indicating the CWD change category and the orange vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

fig.cwd_initdelt <- fig_nums(name = "fig_cwdmeans", 
                          caption = paste0("Mean species plot values for the climatic water deficit initial and change values."))

fig.diff_grow <- fig_nums(name = "fig_diffgrow", 
                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for Figure ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are conifers; those below are hardwoods."))

fig.diff_mort <- fig_nums(name = "fig_diffmort", 
                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for Figure ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled. Species above the dashed line are conifers; those below are hardwoods."))

@@

Western North American forests are experiencing accelerating tree mortality rates in response to intensifying drought and water stress driven by climate change.  Tree mortality and growth reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).  Relating design-based estimates for mortality and growth to a climate variable can be challenging, as directly assessing mortality and growth requires the visitation of plots at two different points in time.  Plots can differ both in their initial climate variable value and the change in that value between visits, necessitating estimation domains that jointly consider initial conditions and temporal change. We developed an analysis framework that partitions plots into domains defined by their initial-visit climatic water deficit (CWD) value and the change in the CWD value between visits.  We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r n.spp` species over decade-long intervals.  We estimated decadal growth and mortality overall, by state, and across six CWD domains per species, along with associated measures of uncertainty.  Species responses were highly variable, with no universal pattern of increased mortality or reduced growth in domains experiencing elevated water stress. Several species exhibited overall mortality rates exceeding 0.20 (i.e., 20% decadal mortality), with estimates of 0.34 for *Picea engelmannii* in Washington, 0.27 for *Acer macrophyllum* in California, and 0.37 for *Pinus contorta* in Washington. Domain-specific patterns aligned with known regional disturbances (bark beetles, drought, fire) for multiple species. Our study contributes a template for monitoring forest vital rates as ecosystem-level indicators of climate change response, and our species-specific findings identify or validate populations warranting immediate conservation assessment that can inform forest management strategies.

@@

Carbon starvation and hydraulic failure can drive drought-induced mortality (see reviews by Adams et al. 2010, Sala et al. 2010, 2012, and McDowell et al. 2008).  Carbon starvation occurs when isohydric species (e.g., piñon pine, *Pinus edulus*; Breshears et al. 2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.  Anisohydric plants, or those that do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al. 2015). Adams et al. (2017) performed a review of drought-induced mortality across 26 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.  Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al. 2022).

@@

Data from the U.S. National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth. Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced sampling design (McRoberts 2005, Reams et al. 2005, Woudenberg et al. 2010).  The sampling design allows for approximately unbiased design-based inference of population characteristics across large geographical regions.  Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.

Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates. Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of FIA plot temperatures for seedlings from those of trees. However, FIA plot revisit data were mostly incomplete at the time of their analysis. In contrast, Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.  Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots that gained trees of a given species and plots that lost trees.

@@

Our study has two objectives.  The first is to demonstrate, across 23 species, a design-based estimation approach for tree growth and mortality that simultaneously accounts for initial climatic conditions and change between measurements.  This addresses a gap in design-based analyses of forest change.  Second, we provide current growth and mortality estimates for our analysis species across our study area and by state to establish species status and to inform our climate analysis findings.

@@

Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.  Forestland is defined as an area of at least 4,050 m$^2$ (i.e., one acre) that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al. 2005).  We constrained our analysis to only include remeasurements for areas considered forestland during both visits.

In the coterminous United States, the FIA program uses a spatially balanced, quasi-systematic sampling design. Plots are randomly located and produce sampling intensities of approximately 2,400 hectares per plot, with some denser areas such as those located on national forest system lands. In the western United States, each plot is assigned to one of ten panels, and approximately one panel is measured each year in perpetuity, yielding ten year remeasurement intervals on each plot, barring exceptions owing to optional temporal intensifications (Thompson 2015). In this study, we use only those plots that are part of the standard sampling intensity, and have been remeasured at an interval of ten years. Each FIA plot consists, in part, of four 7.32 m radius subplots where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.  We did not include data from microplots, where seedlings and saplings are measured, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees. Reams et al. (2005) and Bechtold and Scott (2005) provide plot design details and measurement protocols.

Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure. Tree growth is the observed increase in marked tree bole diameter between visits (cm$^2$) while mortality is captured from the change in tree condition. We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).  The database included plot data sampled from 2001 to 2023.

Remeasured trees within plots were retained in the analysis if they were alive at the initial visit and met the stem diameter criterion during that visit (i.e., plot ingrowth was not included).  The growth analysis additionally restricted trees considered to those that were alive at both visits.  The mortality analysis excluded trees killed through silvicultural practices.  Both analyses excluded trees that shrank in diameter, physically moved due to landslides or other events, were erroneously tallied at either visit, were missed due to data collection procedural changes, or existed at either visit in a nonsampled area.

@@

We obtained plot-level CWD data from the TerraClimate dataset (Abatzoglou et al. 2018) for June, July, and August (summer months) between 1980 and 2023. B. Frank determined that the Pearson's correlation between the fuzzed and unfuzzed CWD plot values was > 0.99, so for simplicity we used the fuzzed CWD values and spatial coordinates. We defined 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year (initial CWD = mean(summer CWD from visit year$_{-19}$ to year$_0$)).  We then found the mean of summer CWD values between visits and including the second visit year (second visit CWD = mean(summer CWD from year$_1$ to year$_10$)).  We then defined 'delta CWD' as the second visit CWD values minus initial CWD values.  Figure `r fig_nums("fig_cwdmap", display = "num")` presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.

## Species  Classification
